---
title: "Project Proposal"
author: "Alex Brown, David Geyfman, Faraz Farooq, Jason Freeberg"
date: "28 October 2016"
output: pdf_document
highlight: espresso
---

# Question Goals
  Our team would like to investigate how a couple's race and years of education relate to the time until a couple's divorce. The full explanations of all covariates are listed below. We will begin with the simplest model solely based on time and event status. Then we will add covariates to the model, check their significance and arrive at a final model.

# Additional Component
  Since our group has one extra person we will take an extra step beyond the base project requirements. After we meet the base requirements on Gauchospace, our group will write a step-by-step tutorial for the project. We will explain the basic statistical methods and functions of the Survival package. We will then publsh the tutorial on Inertia7.com, a website crated by UCSB graduates for hosting data science and statistics project tutorials. 

# Explanation of variables:
### 1) *id*... A couple's identification number during the study

### 2) *edu*... The husband's education level. 
     - 0 = less than 12 years (only high school)
     - 1 = 12 to 15 years (only bachelors or equivalent)
     - 2 = 16 or more years (some form of graduate studies)
### 3) *hblack*... Coded 1 if the husband is black, 0 otherwise
### 4) *mixed*... Coded 1 if the couple is of different ethnicities, 0 otherwise
        - ethnicities are defined as either being "black" or "other"
### 5) *years*... The number of years from wedding to either divorce or censorship
### 6) *div*... The failure indicator
      - 1 = divorced
      - 0 = censored

```{r, warning=FALSE, error=FALSE}
library(ggplot2)
library(survival)
library(cowplot)

# setwd("~/Documents/Fall/PSTAT175/SurvivalAnalysis")

# Load data
colNames <- c("id", "edu", "hblack", "mixed", "years", "div")
divorce <- read.table(file = "divorce.txt", header = F, col.names = colNames)
head(divorce)
```

```{r, echo=FALSE}
# THIS BLOCK WILL NOT BE IN THE PDF FILE, REMEMBER TO GIVE CREDIT

# Define the neatPrint function, makes for nicer output in Rmarkdown
neatPrint <- function(listOfStrings){
  cat(listOfStrings, fill = 1)
}

# Define the ggsurv function

ggsurv <- function(s, CI = 'def', plot.cens = T, surv.col = 'gg.def',
                   cens.col = 'red', lty.est = 1, lty.ci = 2,
                   cens.shape = 3, back.white = F, xlab = 'Time',
                   ylab = 'Survival', main = ''){
 
  library(ggplot2)
  strata <- ifelse(is.null(s$strata) ==T, 1, length(s$strata))
  stopifnot(length(surv.col) == 1 | length(surv.col) == strata)
  stopifnot(length(lty.est) == 1 | length(lty.est) == strata)
 
  ggsurv.s <- function(s, CI = 'def', plot.cens = T, surv.col = 'gg.def',
                       cens.col = 'red', lty.est = 1, lty.ci = 2,
                       cens.shape = 3, back.white = F, xlab = 'Time',
                       ylab = 'Survival', main = ''){
 
    dat <- data.frame(time = c(0, s$time),
                      surv = c(1, s$surv),
                      up = c(1, s$upper),
                      low = c(1, s$lower),
                      cens = c(0, s$n.censor))
    dat.cens <- subset(dat, cens != 0)
 
    col <- ifelse(surv.col == 'gg.def', 'black', surv.col)
 
    pl <- ggplot(dat, aes(x = time, y = surv)) +
      xlab(xlab) + ylab(ylab) + ggtitle(main) +
      geom_step(col = col, lty = lty.est)
 
    pl <- if(CI == T | CI == 'def') {
      pl + geom_step(aes(y = up), color = col, lty = lty.ci) +
        geom_step(aes(y = low), color = col, lty = lty.ci)
    } else (pl)
 
    pl <- if(plot.cens == T & length(dat.cens) > 0){
      pl + geom_point(data = dat.cens, aes(y = surv), shape = cens.shape,
                       col = cens.col)
    } else if (plot.cens == T & length(dat.cens) == 0){
      stop ('There are no censored observations')
    } else(pl)
 
    pl <- if(back.white == T) {pl + theme_bw()
    } else (pl)
    pl
  }
 
  ggsurv.m <- function(s, CI = 'def', plot.cens = T, surv.col = 'gg.def',
                       cens.col = 'red', lty.est = 1, lty.ci = 2,
                       cens.shape = 3, back.white = F, xlab = 'Time',
                       ylab = 'Survival', main = '') {
    n <- s$strata
 
    groups <- factor(unlist(strsplit(names
                                     (s$strata), '='))[seq(2, 2*strata, by = 2)])
    gr.name <-  unlist(strsplit(names(s$strata), '='))[1]
    gr.df <- vector('list', strata)
    ind <- vector('list', strata)
    n.ind <- c(0,n); n.ind <- cumsum(n.ind)
    for(i in 1:strata) ind[[i]] <- (n.ind[i]+1):n.ind[i+1]
 
    for(i in 1:strata){
      gr.df[[i]] <- data.frame(
        time = c(0, s$time[ ind[[i]] ]),
        surv = c(1, s$surv[ ind[[i]] ]),
        up = c(1, s$upper[ ind[[i]] ]),
        low = c(1, s$lower[ ind[[i]] ]),
        cens = c(0, s$n.censor[ ind[[i]] ]),
        group = rep(groups[i], n[i] + 1))
    }
 
    dat <- do.call(rbind, gr.df)
    dat.cens <- subset(dat, cens != 0)
 
    pl <- ggplot(dat, aes(x = time, y = surv, group = group)) +
      xlab(xlab) + ylab(ylab) + ggtitle(main) +
      geom_step(aes(col = group, lty = group))
 
    col <- if(length(surv.col == 1)){
      scale_colour_manual(name = gr.name, values = rep(surv.col, strata))
    } else{
      scale_colour_manual(name = gr.name, values = surv.col)
    }
 
    pl <- if(surv.col[1] != 'gg.def'){
      pl + col
    } else {pl + scale_colour_discrete(name = gr.name)}
 
    line <- if(length(lty.est) == 1){
      scale_linetype_manual(name = gr.name, values = rep(lty.est, strata))
    } else {scale_linetype_manual(name = gr.name, values = lty.est)}
 
    pl <- pl + line
 
    pl <- if(CI == T) {
      if(length(surv.col) > 1 && length(lty.est) > 1){
        stop('Either surv.col or lty.est should be of length 1 in order
             to plot 95% CI with multiple strata')
      }else if((length(surv.col) > 1 | surv.col == 'gg.def')[1]){
        pl + geom_step(aes(y = up, color = group), lty = lty.ci) +
          geom_step(aes(y = low, color = group), lty = lty.ci)
      } else{pl +  geom_step(aes(y = up, lty = group), col = surv.col) +
               geom_step(aes(y = low,lty = group), col = surv.col)}
    } else {pl}
 
 
    pl <- if(plot.cens == T & length(dat.cens) > 0){
      pl + geom_point(data = dat.cens, aes(y = surv), shape = cens.shape,
                      col = cens.col)
    } else if (plot.cens == T & length(dat.cens) == 0){
      stop ('There are no censored observations')
    } else(pl)
 
    pl <- if(back.white == T) {pl + theme_bw()
    } else (pl)
    pl
  }
  pl <- if(strata == 1) {ggsurv.s(s, CI , plot.cens, surv.col ,
                                  cens.col, lty.est, lty.ci,
                                  cens.shape, back.white, xlab,
                                  ylab, main)
  } else {ggsurv.m(s, CI, plot.cens, surv.col ,
                   cens.col, lty.est, lty.ci,
                   cens.shape, back.white, xlab,
                   ylab, main)}
  pl
}

```

# Exploratory Analysis

## The plots show the following estimates (clockwise from top left):
### 1) The baseline KM estimate for all observations
### 2) The KM estimates for the level of education
### 3) The KM estimates for mixed and non-mixed couples
### 4) The KM estimates for couples with and without a black husband

# GGSURV : This is like plot.surv but makes plot look much nicer
```{r}
divorcefit <- Surv(time = divorce$years, event = divorce$div)
divorceKM <- survfit(divorcefit ~ 1)
educateKM <- survfit(divorcefit ~ divorce$edu)
mixedKM <- survfit(divorcefit ~ divorce$mixed)
hblackKM <- survfit(divorcefit ~ divorce$hblack)

a <- ggsurv(divorceKM) +
  ggtitle("Baseline Model") +
  ylab("Survival Probability") +
  xlab("Time in Years")

b <- ggsurv(educateKM, plot.cens = F) +
  ggtitle("Estimates by Education Level") +
  ylab("Survival Probability") +
  xlab("Time in Years")

c <- ggsurv(mixedKM, plot.cens = F) +
  ggtitle("Estimates for Mixed \n and Non-Mixed Couples") +
  ylab("Survival Probability") +
  xlab("Time in Years")

d <- ggsurv(hblackKM, plot.cens = F) +
  ggtitle("Estimates for Black \n and Non-Black Husbands") +
  ylab("Survival Probability") +
  xlab("Time in Years")

plot_grid(a,b,c,d, nrow = 2, ncol = 2)
```

# Cox Proportional Hazards Model
```{r, echo = FALSE}
logRank1 <- coxph(divorcefit ~ 1)
summary(logRank1)
```

# Furthur Exploratory Analysis
```{r}
ggplot(divorce) +
  geom_boxplot(mapping = aes(x = as.factor(edu), y = years, 
                             group = edu, fill = as.factor(edu)))
ggplot(divorce) +
  geom_boxplot(mapping = aes(x = as.factor(hblack), y = years,
                             group = hblack, fill = as.factor(hblack)))
```

```{r, echo = FALSE}
Ncensored <- sum( divorce$div == 0 )
Ndiv <- nrow(divorce) - Ncensored

a <- paste("Number of censored observations =", Ncensored)
b <- paste("Number of observed divorces =", Ndiv)

neatPrint(c(a,b))
```
